Correspondence in Significance

### 2018-06-28 significance/significance.Rmd

Programs of research that examine the replicability of research findings have become more common in the social sciences. 
On its face, assessing replication seems trivial: just repeat an experiment and check that you get the same results.
However analyses of these programs reveal that we might mean several different things when we refer to "the same results."
Indeed, the Open Science Collaboration (OSC) (2015) argued, "There is no single standard for evaluating replication success," and then analyzed replicate pairs in a variety of ways.

Despite immediate criticism of the analyses (see Gilbert, Pettigrew, and King, 2015), the core set of OSC analyses has become something of a standard in replication research.
Camerer, et al. (2016) conducted an almost identical research design with the same analytic methods on findings in experimental economics. 
Schweinsberg, et al. (2016), while conducting several---i.e., more than one---replicates of the same findings, aggregated replicate results so that they could use the same analyses as the OSC.
Thus, the idea that replication is impossible to assess has persisted across multiple high-profile replication research efforts.

Given their prevalence in the replication literature, the analyses conducted by the OSC have a few important aspects worth exploring.
The first involves the definitions of replication that they imply. 
For instance, comparing $p$-values versus effect estimates may suggest very different ways we may think about studies' "results" and whether they are "the same."
The second are the properties of the methods used to assess definitions of replication.
For most replication studies, authors seek to make determinations about whether a finding was successfully replicated.
Doing so amid uncertainty---such as in effect estimates---runs into similar issues of null hypothesis tests. 
Thus, unambiguous interpretation of the results of replication research, can depend on the properties of the methods authors use to make these determinations.

In this paper, we explore methods for assessing replication as proposed by the OSC.
We argue that definitions of replication ought to involve effect parameters, rather than $p$-values. 
Using this logic, we then frame the set of analyses common in replication research as null hypothesis tests about replication. 
The properties of these tests are computed under some simplifying assumptions, and shown that there are conditions where they may be suboptimal.

@@

In the context of replication studies, we have at least $k \geq 2$ such findings, so let $\theta_i$ describe the results of the $i$th study.
For programs such as that of the OSC, this involves $\theta_1$ for the original study and $\theta_2$ for the replicate.
However, we do not actually observe $\theta_1$ and $\theta_2$, and instead we estimate them by $T_1$ and $T_2$, and these estimates have variances $v_1$ and $v_2$.
The estimates and their variances are what get reported in typical statistical analyses of experiments.

@@

We argue that one useful definition of replication involves the similarity not of estimates $T_i$, but of actual true effects $\theta_i$.
The $T_i$ might differ because of both differences in the underlying true effects, but also due to random chance.
The same can be said for the resulting $p$-values---$p_1$ and $p_2$.
The $\theta_i$ might differ, however, because studies sample participants from different populations, or due to differences in contexts and conditions between experiments.
Thus, it may make sense to define replication in terms of differences in experiments rather than statistical noise.
Note that there is a connection to the potential outcomes framework in causal inference; for discussion see Wong and Steiner (2018).

In that vein, $\theta_1 = \theta_2$ would be a convenient starting point for defining replication.
However, there is evidence that even in the face of strong theory and sound scientific practice, that studies may still get slightly different results.
Thus, a more practical definition might posit that studies replicate if $\theta_1 \approx \theta_2$.
For further discussion on this distinction see Hedges and Schauer (2018).
The key distinction here, is that the \textit{definition} of replication involves the parameters $\theta_i$, while \textit{analyses} involve the estimates $T_i$ and their variances $v_i$

For this paper, we interpret the OSC analyses in the tradition of statistical hypothesis testing.
Each analysis can be seen of a test of the null hypothesis $H_0: \theta_1 = \theta_2$, the metric used can be seen as a test statistic, and determinations that a finding failed to replicate correspond to a rejection of this null hypothesis.
Note that this is not the only structure we might infer here, and is not even the only way to operationalize these methods as hypothesis tests.
However, it is in the tradition of statistical hypothesis tests about heterogeneity across effects.

@@

One metric used by the OSC to determine if studies successfully replicated to to check that both the original and replicate study corresponded in sign and statistical significance.
That is, findings were deemed to have replicated if both studies found significant effects in the same direction---i.e., both positive---or if both studies found null effects.
This type of metric would seem to answer an important question about replication, particularly of published research which more often involves statistically significant findings.
Indeed, Fisher (1935) noted that "we may say that a phenomenon is experimentally demonstrable when we know how to conduct an experiment which will rarely fail to give us statistically significant results."

@@

To get a sense of the properties of this test, assume that $\theta_1 = \theta_2 = \theta$, or that the studies replicate exactly.
If this were true, we would want this test to infrequently reject this null hypothesis; in statistical terms, we would want the type I error rate to be low.
The null hypothesis would be rejected if $T_1$ is significant and $T_2$ is either not significant or significant and has an opposite direction (or visa versa). 
Thus, the type I error rate is given by
<!-- \begin{align} -->
<!-- \begin{split} -->
<!-- & P[T_1/\sqrt{v_1} > 1.96]P[T_2/\sqrt{v_2} \leq 1.96] +  -->
<!-- P[T_1/\sqrt{v_1} \leq 1.96]P[T_2/\sqrt{v_2} > 1.96] \\ +  -->
<!-- & P[T_1/\sqrt{v_1} < -1.96]P[T_2/\sqrt{v_2} \geq -1.96] +  -->
<!-- P[T_1/\sqrt{v_1} \geq -1.96]P[T_2/\sqrt{v_2} < -1.96] -->
<!-- \end{split} -->
<!-- \end{align} -->
<!-- Under the model, this can be written as  -->
\begin{align*}
P[\text{Type I Error}]
& = [1 - \Phi\left(1.96 - \theta/\sqrt{v_1}\right)] \Phi\left(1.96 - \theta/\sqrt{v_2}\right)\\ 
&\quad + \Phi\left(1.96 - \theta/\sqrt{v_1}\right) [1 - \Phi\left(1.96 - \theta/\sqrt{v_2}\right)]\\ 
&\quad + \Phi\left(-1.96 - \theta/\sqrt{v_1}\right) [1 - \Phi\left(-1.96 - \theta/\sqrt{v_2}\right)]\\
&\quad + [1 - \Phi\left(-1.96 - \theta/\sqrt{v_1}\right)] \Phi\left(-1.96 - \theta/\sqrt{v_2}\right)
\end{align*}
\noindent
where $\Phi$ is the standard normal distribution function.

@@

In Figure \ref{fig:sig:falsepos}, we see that the false positive rates are nearly all above the nominal 5\%. 
This is not necessarily an artifact of the scale of the graph. 
Indeed, if study 2 had an infinite sample size, so that $v_2 \to 0$ and $\theta_2/\sqrt{v_2} \to 0$, then the false positive rate of this test for replication converges to one minus the power of the initial study:
\[
\text{Error rate } \to \beta_1
\]
where $1 - \beta_1$ is the power of study 1.
Thus, unless study 1 has at least 95\% power, determining if studies replicated according to correspondence of sign and statistical significance can result in type I error rates well above 5\%.

@@

Conversely, if we assume $\theta_1 \neq \theta_2$, then we would want the test to reject the null hypothesis frequently---that is, we would want a low type II error rate.
To conclude that the studies do replicate would require both findings to be significant and in the same direction (e.g., positive), or to both be nonsignificant.
Under the model, the probability of concluding that the studies do replicate when $\theta_1 \neq \theta_2$ can be written as:
\begin{align*}
P[\text{Type II Error}]
& = [1 - \Phi\left(1.96 - \theta_1/\sqrt{v_1}\right)][1 - \Phi\left(1.96 - \theta_2/\sqrt{v_2}\right)]\\ 
& + [\Phi\left(1.96 - \theta_1/\sqrt{v_1}\right) - \Phi\left(-1.96 - \theta_1/\sqrt{v_1}\right)]
    [\Phi\left(1.96 - \theta_2/\sqrt{v_2}\right) - \Phi\left(-1.96 - \theta_2/\sqrt{v_2}\right)]\\ 
& + \Phi\left(-1.96 - \theta_1/\sqrt{v_1}\right) \Phi\left(-1.96 - \theta_2/\sqrt{v_2}\right)
\end{align*}

When one of the effect parameters, say $\theta_2$ is equal to zero, then this can be written in terms of the power of the other study:
\[
P[\text{Type II Error}]
 = 0.025 + 0.925 \beta_1
\]
where $1 - \beta_1$ is the power of study 1. 
Thus, if study 1 has low power, so that $\beta_1$ is large, then this test would be likely to conclude that the studies successfully replicated. 
For instance, if study 1 has power 80\%, then this test would conclude that the studies do not replicate 21\% of the time. 
However, if the power were lower, say 60\%, then the test would make this error 39.5\% of the time.

If both of the results are nonzero but the same direction (e.g., both positive), then the error rate will be large whenever both studies have high (or low) power. 
For instance, if $\theta_1$ is much larger than $\theta_2$, but both have power 80\%, then this test will conclude that the studies replicate almost 68\% of the time! 
Similarly, if both studies have low power then there is a higher probability that both will return nonsignificant results, and hence we would conclude that the studies replicated despite $\theta_1 \neq \theta_2$.
In fact, the only way for the false negative rate to decrease is for one study to have very high power, and the other to have very low power.

>>>>

<!-- Programs of research that examine the replicability of research findings have become more common in the social sciences.  -->
<!-- On its face, assessing replication seems trivial: just repeat an experiment and check that you get the same results. -->
<!-- However analyses of these programs reveal that we might mean several different things when we refer to "the same results." -->
<!-- Indeed, the Open Science Collaboration (RPP) (2015) argued, "There is no single standard for evaluating replication success," and then analyzed replicate pairs in a variety of ways. -->

<!-- Despite immediate criticism of the analyses (see Gilbert, Pettigrew, and King, 2015), the core set of RPP analyses has become something of a standard in replication research. -->
<!-- Camerer, et al. (2016) conducted an almost identical research design with the same analytic methods on findings in experimental economics.  -->
<!-- Schweinsberg, et al. (2016), while conducting several---i.e., more than one---replicates of the same findings, aggregated replicate results so that they could use the same analyses as the RPP. -->
<!-- Thus, the idea that replication is impossible to assess has persisted across multiple high-profile replication research efforts. -->

<!-- Given their prevalence in the replication literature, the analyses conducted by the RPP have a few important aspects worth exploring. -->
<!-- The first involves the definitions of replication that they imply.  -->
<!-- For instance, comparing $p$-values versus effect estimates may suggest very different ways we may think about studies' "results" and whether they are "the same." -->
<!-- The second are the properties of the methods used to assess definitions of replication. -->
<!-- For most replication studies, authors seek to make determinations about whether a finding was successfully replicated. -->
<!-- Doing so amid uncertainty---such as in effect estimates---runs into similar issues of null hypothesis tests.  -->
<!-- Thus, unambiguous interpretation of the results of replication research, can depend on the properties of the methods authors use to make these determinations. -->

<!-- In this paper, we explore methods for assessing replication as proposed by the RPP. -->
<!-- We argue that definitions of replication ought to involve effect parameters, rather than $p$-values.  -->
<!-- Using this logic, we then frame the set of analyses common in replication research as null hypothesis tests about replication.  -->
<!-- The properties of these tests are computed under some simplifying assumptions, and shown that there are conditions where they may be suboptimal. -->

@@

In the context of replication studies, we have at least $k \geq 2$ such findings, so let $\theta_i$ describe the results of the $i$th study.
For programs such as that of the Replication Project: Psychology (RPP) (Open Science Collaboration, 2015), this involves $\theta_1$ for the original study and $\theta_2$ for the replicate.
However, we do not actually observe $\theta_1$ and $\theta_2$, and instead we estimate them by $T_1$ and $T_2$, and these estimates have variances $v_1$ and $v_2$.
The estimates and their variances are what get reported in typical statistical analyses of experiments.

@@

We argue that one useful definition of replication involves the similarity not of estimates $T_i$, but of actual true effects $\theta_i$.
The $T_i$ might differ because of both differences in the underlying true effects, but also due to random chance.
The same can be said for the resulting $p$-values---$p_1$ and $p_2$.
The $\theta_i$ might differ, however, because studies sample participants from different populations, or due to differences in contexts and conditions between experiments.
Thus, it may make sense to define replication in terms of differences in experiments rather than statistical noise.
Thus, we offer $\theta_1 = \theta_2$ as a definition of replication worth testing.
<!-- Note that there is a connection to the potential outcomes framework in causal inference; for discussion see Wong and Steiner (2018). -->

<!-- In that vein, $\theta_1 = \theta_2$ would be a convenient starting point for defining replication. -->
<!-- However, there is evidence that even in the face of strong theory and sound scientific practice, that studies may still get slightly different results. -->
<!-- Thus, a more practical definition might posit that studies replicate if $\theta_1 \approx \theta_2$. -->
<!-- For further discussion on this distinction see Hedges and Schauer (2018). -->
<!-- The key distinction here, is that the \textit{definition} of replication involves the parameters $\theta_i$, while \textit{analyses} involve the estimates $T_i$ and their variances $v_i$ -->

<!-- For this paper, we interpret the RPP analyses in the tradition of statistical hypothesis testing. -->
<!-- Each analysis can be seen of a test of the null hypothesis $H_0: \theta_1 = \theta_2$, the metric used can be seen as a test statistic, and determinations that a finding failed to replicate correspond to a rejection of this null hypothesis. -->
<!-- Note that this is not the only structure we might infer here, and is not even the only way to operationalize these methods as hypothesis tests. -->
<!-- However, it is in the tradition of statistical hypothesis tests about heterogeneity across effects. -->

@@

One metric used by the RPP to determine if studies successfully replicated was if the original and replicate study corresponded in sign and statistical significance.
Findings were deemed to have replicated if both studies found significant effects in the same direction---i.e., both positive---or if both studies found null effects.
Heuristically, this would seem logical, and indeed, Fisher (1935) noted that "we may say that a phenomenon is experimentally demonstrable when we know how to conduct an experiment which will rarely fail to give us statistically significant results."

@@

To get a sense of the properties of this test, assume that $\theta_1 = \theta_2 = \theta$, or that the studies replicate exactly.
If this were true, we would want this test to infrequently reject this null hypothesis; in statistical terms, we would want the type I error rate to be low.
The null hypothesis would be rejected if $T_1$ is significant and $T_2$ is either not significant or significant and has an opposite direction (or visa versa). 
If both $T_1$ and $T_2$ are subject to two-tailed $\alpha = 0.05$ level tests, the type I error rate is given by
<!-- \begin{align} -->
<!-- \begin{split} -->
<!-- & P[T_1/\sqrt{v_1} > 1.96]P[T_2/\sqrt{v_2} \leq 1.96] +  -->
<!-- P[T_1/\sqrt{v_1} \leq 1.96]P[T_2/\sqrt{v_2} > 1.96] \\ +  -->
<!-- & P[T_1/\sqrt{v_1} < -1.96]P[T_2/\sqrt{v_2} \geq -1.96] +  -->
<!-- P[T_1/\sqrt{v_1} \geq -1.96]P[T_2/\sqrt{v_2} < -1.96] -->
<!-- \end{split} -->
<!-- \end{align} -->
<!-- Under the model, this can be written as  -->
\begin{align*}
P[\text{Type I Error}]
& = [1 - \Phi\left(1.96 - \theta/\sqrt{v_1}\right)] \Phi\left(1.96 - \theta/\sqrt{v_2}\right)\\ 
&\quad + \Phi\left(1.96 - \theta/\sqrt{v_1}\right) [1 - \Phi\left(1.96 - \theta/\sqrt{v_2}\right)]\\ 
&\quad + \Phi\left(-1.96 - \theta/\sqrt{v_1}\right) [1 - \Phi\left(-1.96 - \theta/\sqrt{v_2}\right)]\\
&\quad + [1 - \Phi\left(-1.96 - \theta/\sqrt{v_1}\right)] \Phi\left(-1.96 - \theta/\sqrt{v_2}\right)
\end{align*}
\noindent
where $\Phi$ is the standard normal distribution function.

@@

In Figure \ref{fig:sig:falsepos}, we see that the error rate is largest when the power of both studies is moderate. 
If both studies have high power, then it becomes more likely that both estimates will be significant; and if both studies have low power, then it becomes more likely that both estimates will be nonsignificant. 
However, the false positive rates in Figure \ref{fig:sig:falsepos} are nearly all above the nominal 5\%. 
This is not necessarily an artifact of the scale of the graph. 
Indeed, if study 2 had an infinite sample size, so that $v_2 \to 0$ and $\theta_2/\sqrt{v_2} \to 0$, then the false positive rate of this test for replication converges to one minus the power of the initial study:
\[
\text{Error rate } \to \beta_1
\]
where $1 - \beta_1$ is the power of study 1.
Thus, unless study 1 has at least 95\% power, determining if studies replicated according to correspondence of sign and statistical significance can result in type I error rates well above 5\%.

For reference, suppose all 100 of the RPP studies involved studies that replicated exactly, with effects that are nonzero ($\theta_1 = \theta_2 \neq 0$).
If all studies were powered at 80\%, then we would expect this procedure to falsely determine that 41 findings failed to replicate. 
If the power of the studies was lower, say 60%, then we would expect this procedure to falsely conclude that 48 of findings failed to replicate.

@@

Conversely, if we assume $\theta_1 \neq \theta_2$, then we would want the procedure to determine that the findings do not replicate frequently---that is, we would want a low type II error rate.
To conclude that the studies do replicate would require both findings to be significant and in the same direction (e.g., positive), or to both be nonsignificant.
Under the model, the probability of concluding that the studies do replicate when $\theta_1 \neq \theta_2$ can be written as:
\begin{align*}
P[\text{Type II Error}]
& = [1 - \Phi\left(1.96 - \theta_1/\sqrt{v_1}\right)][1 - \Phi\left(1.96 - \theta_2/\sqrt{v_2}\right)]\\ 
& + [\Phi\left(1.96 - \theta_1/\sqrt{v_1}\right) - \Phi\left(-1.96 - \theta_1/\sqrt{v_1}\right)]
    [\Phi\left(1.96 - \theta_2/\sqrt{v_2}\right) - \Phi\left(-1.96 - \theta_2/\sqrt{v_2}\right)]\\ 
& + \Phi\left(-1.96 - \theta_1/\sqrt{v_1}\right) \Phi\left(-1.96 - \theta_2/\sqrt{v_2}\right)
\end{align*}

When one of the effect parameters, say $\theta_2$ is equal to zero, then this can be written in terms of the power of the other study:
\[
P[\text{Type II Error}]
 = 0.025 + 0.925 \beta_1
\]
where $1 - \beta_1$ is the power of study 1. 
Thus, if study 1 has low power, so that $\beta_1$ is large, then this test would be likely to conclude that the studies successfully replicated. 
For instance, if study 1 has power 80\%, then this test would conclude that the studies do  replicate 21\% of the time. 
However, if the power were lower, say 60\%, then the test would make this error 39.5\% of the time.

If both of the parameters are nonzero but the same direction (e.g., both positive), then the error rate will be large whenever both studies have high (or low) power. 
For instance, if $\theta_1$ is much larger than $\theta_2$, but both have power 80\%, then this test will conclude that the studies replicate almost 68\% of the time! 
Similarly, if both studies have low power then there is a higher probability that both will return nonsignificant results, and hence we would conclude that the studies replicated despite $\theta_1 \neq \theta_2$.
In fact, the only way for the false negative rate to decrease is for one study to have very high power, and the other to have very low power.
